---
title: "Visualization and EDA"
author: "Anyu Zhu"
date: "10/5/2021"
output: github_document
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(ggridges)
library(hexbin)
library(patchwork)
library(haven)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

# theme_set(theme_minimal() + theme(legend.position = "bottom"))

#options(
#  ggplot2.continuous.colour = "viridis",
#  ggplot2.continuous.fill = "viridis"
#)

#scale_colour_discrete = scale_colour_viridis_d
#scale_fill_discrete = scale_fill_viridis_d
```

Load a dataset
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())
```

## Scatter Plot

tmax vs tmin
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()
```
can save ggplots
```{r}
ggp_tmax_tmin = weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point()

ggp_tmax_tmin
# ggsave(ggp_tmax_tmin)
```

## Fancy it up

Add colors, lines, etc,
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + # color globally
  geom_point(alpha = 0.3) + # transparency of points
  geom_smooth(se = FALSE)
```
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = 0.3) + # color only to points
  geom_smooth(se = FALSE)
```
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```
One more scatter plot
```{r}
weather_df %>% 
  ggplot(aes(x = date, y = tmax)) +
  geom_point(alpha = 0.3) +
  facet_grid(. ~ name) +
  geom_smooth(se = FALSE) # se = TRUE give confidence interval
```

## Use data manipulation as part of this

```{r}
weather_df %>% 
  filter(name == "CentralPark_NY") %>% 
  mutate(tmax = tmax * (9/5) + 32,
         tmin = tmin * (9/5) + 32) %>% 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()
```

## Stacking geoms

```{r}
weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_smooth()
```
```{r}
ggplot(weather_df, aes(x = tmax, y = tmin)) + 
  geom_hex()
```

## Univariate Plots

```{r}
weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +
  geom_histogram() + 
  facet_grid(. ~ name)
```
```{r}
weather_df %>% 
  ggplot(aes(x = tmax, fill = name)) +
  geom_density(alpha = 0.3)
```

Box plot
```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_boxplot()
```

Violin plot

```{r}
weather_df %>% 
  ggplot(aes(x = name, y = tmax)) + 
  geom_violin()
```

Without ridges
```{r}
weather_df %>% 
  ggplot(aes(x = tmax, y = name)) + 
  geom_density_ridges(alpha = 0.8, scale = 0.8)
```

## Embedding plots

```{r, fig.width=12}
weather_plot = ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) 
weather_plot
```


## Start with a familiar one

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Temperature at three stations",
    x = "Minimum daily temp (C)",
    y = "Maximum daily temp (C)",
    caption = "Data from package rnoaa with three stations" 
  )
```

## Scales

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Temperature at three stations",
    x = "Minimum daily temp (C)",
    y = "Maximum daily temp (C)",
    caption = "Data from package rnoaa with three stations" 
  ) +
  scale_x_continuous(
    breaks = c(-15, 0, 15),
    labels = c("-15 C", "0", "15")
  ) +
  scale_y_continuous(
    trans = "sqrt",
    position = "right" # label on the right
  )
```

Color Scales
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Temperature at three stations",
    x = "Minimum daily temp (C)",
    y = "Maximum daily temp (C)",
    caption = "Data from package rnoaa with three stations" 
  ) +
  scale_color_hue(
    name = "Location", # change legend title
    h = c(100, 300)) +
  scale_color_viridis_d()
# https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
```

## Themes

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Temperature at three stations",
    x = "Minimum daily temp (C)",
    y = "Maximum daily temp (C)",
    caption = "Data from package rnoaa with three stations" 
  ) +
  scale_color_hue(
    name = "Location", # change label name
    h = c(100, 300)) +
  scale_color_viridis_d() +
  theme_minimal() + # background; do this first!! then update other themes
  theme(legend.position = "bottom")
```
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Temperature at three stations",
    x = "Minimum daily temp (C)",
    y = "Maximum daily temp (C)",
    caption = "Data from package rnoaa with three stations" 
  ) +
  scale_color_hue(
    name = "Location", # change label name
    h = c(100, 300)) +
  scale_color_viridis_d() + 
  ggthemes::theme_excel() +
  theme(legend.position = "bottom")
```

## data in geoms

```{r}
central_park = 
  weather_df %>% 
  filter(name == "CentralPark_NY")

waikiki = 
  weather_df %>% 
  filter(name == "Waikiki_HA")

weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point()



waikiki %>% 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point() + 
  geom_line(data = central_park)
```

## `patchwork`

```{r}
ggp_tmax_tmin = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3)

ggp_prcp_dens = 
  weather_df %>% 
  filter(prcp > 0) %>% 
  ggplot(aes(x = prcp, fill = name)) +
  geom_density(alpha = 0.3)

ggp_tmax_date = 
  weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point() +
  geom_smooth(se = FALSE)

(ggp_tmax_tmin + ggp_prcp_dens) / (ggp_tmax_date) # patchwork
```

## data manipulation

examples on factor
```{r}
weather_df %>% 
  mutate(
    name = fct_reorder(name, tmax) # order according to values; default: alphabetic
  ) %>% 
  ggplot(aes(x = name, y = tmax, fill = name)) +
    geom_boxplot()
```

About tmax and tmin

```{r}
weather_df %>% 
  pivot_longer(
    tmax:tmin,
    names_to = "observations",
    values_to = "temperature"
  ) %>% 
  ggplot(aes(x = temperature, fill = observations)) +
  geom_density(alpha = 0.3) +
  facet_grid(. ~ name)
```

```{r}
pulse_df = haven::read_sas("data/public_pulse_data.sas7bdat") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    values_to = "bdi",
    names_prefix = "bdi_score_"
  ) %>% 
  mutate(visit = recode(visit, "bl" = "00m"))

pulse_df %>% 
  ggplot(aes(x = visit, y = bdi)) +
  geom_boxplot()
```

Longitudinal analysis

```{r}
pulse_df %>% 
  ggplot(aes(x = visit, y = bdi))+
  geom_point() + 
  geom_line(aes(group = id), alpha = 0.3) # keep track of each person
```



```{r}
weather_df =  
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())
```

## `group_by`
```{r}
weather_df %>% 
  group_by(name, month)
```
## count some things
```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n()) # give group size

weather_df %>% 
  count(name)
```

`table` doesn't produce a data.frame, while summarize does. NOT use table()
```{r}
weather_df %>% 
  pull(name) %>% 
  table()
```

```{r}
weather_df %>% 
  janitor::tabyl(month, name) # two interested variable
```

## More general summaries

count, take means, look at SDs

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    n_obs = n(),
    mean_tmax = mean(tmax, na.rm = TRUE), # remove NA values
    median_tmax = median(tmax, na.rm = TRUE),
    sd_tmax = sd(tmax, na.rm = TRUE)
  ) %>% 
ggplot(aes(x = month, y = mean_tmax, color = name)) + 
  geom_point() +
  geom_line()
```

Formatting table output
```{r}
weather_df %>% 
  group_by(name) %>% 
  summarize(n_obs = n(),
            mean_tmax = mean(tmax, na.rm = TRUE)) %>% 
  knitr::kable(digits = 2)
```

## grouped `mutate`

```{r}
weather_df %>% 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax
      ) %>% 
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()
```

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    tmax_rank = min_rank(desc(tmax))
  ) %>% 
  filter(tmax_rank < 2)
```

Lagged variables

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    lagged_tmax = lag(tmax, n = 1), # look at values a day before (n = 1)
    tmax_diff = tmax - lagged_tmax
  ) %>% 
  summarize(diff_sd = sd(tmax_diff, na.rm = TRUE))
```

## Limitations

What if "summary" is a linear model?

```{r, eval = FALSE}
weather_df %>% 
  group_by(name) %>% 
  summarize(cor_tmin_tmax = cor(tmin, tmax, use = "complete"))

weather_df %>% 
  filter(name == "CentralPark_NY") %>% 
  lm(tmax ~ tmin, data = .)
```


























